\morus: Biases in the Stream

### 2018-05-09 morusAC_01_Introduction.tex
...
Conclusion

Authenticated Encryption (AE) schemes combine the functionality of symmetric encryption schemes and message authentication codes.
Based on a shared secret key $K$, they encrypt a plaintext message $M$ to a ciphertext $C$ and authentication tag $T$ in order to protect both the confidentiality and the authenticity of $M$.
Most modern authenticated encryption algorithms are nonce-based schemes with associated data (AEAD),
where $(C, T)$ additionally depends on a unique nonce $N$ (or initialization value IV) and optional associated metadata $A$.
One of the most prominent standardized AEAD designs is \cipher{AES-GCM} \cite{indocryptMcGrewV04,nistDworkin07},
which is widely deployed in protocols such as \cipher{TLS v1.3}.

@@

% Paragraph: Details on keystream bias
Our main result is a linear approximation \cite{eurocryptMatsui93,eurocryptMatsuiY92} linking ciphertext bits of five consecutive encryption blocks.
%As a consequence of this trail, we exhibit a biased linear combination of ciphertext bits, which holds across any given sequence of five consecutive encryption steps.
%Technically, this assumes that the inner state of the cipher at the start of the trail is uniformly random, but this is a fair assumption since the inner state should be indistinguishable from uniform randomness by design.
%Moreover, that is the only assumption required; in particular, the secret key or the nonce used to initialize the inner state of MORUS are irrelevant: the bias is independent of their value.
More precisely, we show that when encrypting a fixed plaintext block multiple times, a linear correlation exists between some output bits of the cipher.
Moreover, the bias depends purely on the plaintext, and not on the secret key of the cipher.
In principle, this property could be used to recover unknown bits of a plaintext encrypted a large number of times, provided an initial segment of the plaintext is known.
For \morus[1280], the \morus family member with 256-bit security, the linear bias is $2^{-73}$
\todo{double-check bias vs. correlation everywhere}
and can be exploited using about $2^{152}$ encrypted blocks.
\todo{double-check}
We note that rekeying does not prevent the attack: the biases are independent of the secret encryption key, and can be exploited as long as a given plaintext block is encrypted sufficiently often, regardless of whether each encryption uses a different key.

@@

\subsubsection*{Paper Outline.}
This paper is organized as follows.
We first provide a brief description of \morus in \autoref{sec/Preliminaries}.
In \autoref{sec/introminimorus}, we introduce \minimorus, an abstraction of \morus based on a certain class of rotational symmetries.
We analyze this simplified scheme in \autoref{sec/minitrails} and provide a ciphertext-only linear approximation with a weight of 16.
We then extend our result to the full scheme in \autoref{sec/fulltrails}, showing a bias in the keystream over 5 rounds,
and discuss the implications of our observation for the security of \morus in \autoref{sec/Discussion}.
In \autoref{sec/IniFin}, we present our results on the security of \morus with round-reduced initialization (in a nonce-misuse setting) or finalization.
Finally, we conclude in \autoref{TODO}.

>>>>

Authenticated Encryption (AE) schemes combine the functionality of symmetric encryption schemes and message authentication codes.
Based on a shared secret key $K$, they encrypt a plaintext message $M$ to a ciphertext $C$ and authentication tag $T$ in order to protect both the confidentiality and the authenticity of $M$.
Most modern authenticated encryption algorithms are nonce-based schemes with associated data (AEAD),
where $(C, T)$ additionally depends on a unique nonce $N$ (or initialization value IV) and optional associated metadata $A$.
One of the most prominent standardized AEAD designs is \cipher{AES-GCM} \cite{indocryptMcGrewV04,nistDworkin07},
which is widely deployed in protocols such as \cipher{TLS} (since v1.2).

@@

% Paragraph: Details on keystream bias
Our main result is a linear approximation \cite{eurocryptMatsui93,eurocryptMatsuiY92} linking ciphertext bits of five consecutive encryption blocks.
%As a consequence of this trail, we exhibit a biased linear combination of ciphertext bits, which holds across any given sequence of five consecutive encryption steps.
%Technically, this assumes that the inner state of the cipher at the start of the trail is uniformly random, but this is a fair assumption since the inner state should be indistinguishable from uniform randomness by design.
%Moreover, that is the only assumption required; in particular, the secret key or the nonce used to initialize the inner state of MORUS are irrelevant: the bias is independent of their value.
More precisely, we show that when encrypting a fixed plaintext block multiple times, a linear correlation exists between some output bits of the cipher.
Moreover, the bias depends purely on the plaintext, and not on the secret key of the cipher.
In principle, this property could be used to recover unknown bits of a plaintext encrypted a large number of times, provided an initial segment of the plaintext is known.
For \morus[1280] with 256-bit keys, the linear bias is $2^{-73}$
\todo{double-check bias vs. correlation everywhere}
and can be exploited using about $2^{152}$ encrypted blocks.
\todo{double-check}
We note that rekeying does not prevent the attack: the biases are independent of the secret encryption key, and can be exploited as long as a given plaintext block is encrypted sufficiently often, regardless of whether each encryption uses a different key.

@@

\subsubsection*{Paper Outline.}
This paper is organized as follows.
We first provide a brief description of \morus in \autoref{sec/Preliminaries}.
In \autoref{sec/introminimorus}, we introduce \minimorus, an abstraction of \morus based on a certain class of rotational symmetries.
We analyze this simplified scheme in \autoref{sec/minitrails} and provide a ciphertext-only linear approximation with a weight of 16.
We then extend our result to the full scheme in \autoref{sec/fulltrails}, showing a bias in the keystream over 5 rounds,
and discuss the implications of our observation for the security of \morus in \autoref{sec/Discussion}.
In \autoref{sec/IniFin}, we present our results on the security of \morus with round-reduced initialization (in a nonce-misuse setting) or finalization.
Finally, we conclude in \autoref{sec/Conclusion}.
